Response Letter for Manuscript ID BIOM2020344P - 'Efficient Detection and Classification of Epigenomic Changes Under Multiple Conditions

### 2021-03-02 Latex/letter/letter.Rmd
Update letter.Rmd
typo fixes and added comments to be addressed

> We thank the reviewer for this constructive comment. We cite in our paper the benchmarking article 'A comprehensive comparison of tools for differential ChIP-seq analysis' (Briefings in Bioinformatics, 2016). However, we authors are not aware of a more recent comprehensive review of DPCs algorithms in the literature. As suggested we add the following sentence to our real data analysis section 'We compared the genome-wide performance of the presented model with the widely used DPCs ChIPComp, csaw, DiffBind, diffReps, RSEG, and THOR, a list that covers a variety of algorithmic approaches for both broad and short data sets.' We believe that such a list covers the most used algorithms for DPC in thie field of bioinformatics.

@@

> As mentioned above, our method is able to account go GC-bias via model offsets. These model offsets can be obtained by methods such as the one proposed by Teng and Irizarry (2017). We present a simulation study in Section B1.3 of the Web Appendix showing that epigraHMM leads to higher sensitivity/specificity when accounting for GC-bias even after read count normalization using the count-normalizing denominators from Teng and Irizarry (2017). However, for the broad marks that we analized in our manuscript, we did not observe a substantial effect of GC-content on read count distribution and did not observe any substantial improvement in accounting for GC-content in our model via offsets when utilizing the method from Teng and Irizarry (2017), a method that was developed only for narrow marks characterized by sharp peaks. In fact, the authors do not recommend the use of their method when analyze epigenomic marks with broad enrichment. We did not explore alternative methods for GC-bias correction for broad marks as we believe this task is beyond the scope of our work.

@@

> In the main text, we refer the reader to Section B5 of the Web Appendix. To estimate the model coefficients during the $t^{th}$ iteration of the EM-algorithm, one needs to find the optimal values $\beta_{1}^{(t)}$, $\beta_{3}^{(t)}$, $\lambda_{1}^{(t)}$, and $\lambda_{3}^{(t)}$ that maximize the quantity $Q_{1}\left(\boldsymbol{\psi}_{1}|\boldsymbol{\Psi}^{(t)}\right) + Q_{2}\left(\boldsymbol{\delta},\boldsymbol{\psi}_{2}|\boldsymbol{\Psi}^{(t)}\right) + Q_{3}\left(\boldsymbol{\psi}_{3}|\boldsymbol{\Psi}^{(t)}\right)$. This problem is analogous to estimating parameters in a weighted NB regression model, since $Q_{1}$, $Q_{2}$, and $Q_{3}$ can be seen as weighted log-likelihood functions from a NB generalized linear model with means and dispersions parametrized as described in the Section 3.1 of the main article. Exact forms of the $Q$-functions are presented in Equation B.4 of Web Appendix B5.

@@

> Results presented in Table 1 pretain to the full model with all possible combinatorial patterns. This is the 'optimal' model in the sense that the simulated data contains all possible combinatorial patterns of enrichment and background among the condition. The point of Figure 1 is to show that our efficient estimation method via fast RCEM leads to unbiased parameter estimates. Together with Figure 2, Table 1 raises the point that increasing the number of technical or biological replicates leads to better results in terms of sensitivity/specificity of peak calls as well as in the sense of asymptotic efficiency of parameter estimates. We leave the discussion of model selection via BIC in our model to Section B2 of the Web Appendix where we present a simulation study.

@@

> We thank the reviewer for point this out. We believe that comparing the relative bias of parameter estimates across different SNR levels is difficult because different SNRs are associated with different parameter values (some larger than others) in our simulation study. Although we did observe an improvement in performance in terms of model sensitivity/specificity of peak calls in higher SNR data sets, the main driving quantity for the reduction of parameter bias is the number of replicates included in our model. For this reason, we rephrase the mentioned sentence in our paper as "Depending on the number of conditions, the observed relative bias and the range of the reported percentiles tended to decrease as more replicates were included in the analyses. This effect was particularly significant in scenarios with four conditions with respect to parameters $\beta_{3}$ and $\lambda_{3}$. In scenarios with higher number of conditions, these results highlight the importance of experimental replicates to achieve precise parameter estimates.". The estimated SNR level from ENCODE was 3.21 (ratio of mean counts between peak and background regions) for H3K37me3. This is now reported in our manuscript. We present the estimate SNR value for other scenarios in our Web Appendix (Web Tables 2-3)

@@

> We believe our method is able to handle such situations. Note, for instance, in Figure 4D of our manuscript that we report differential peak calls from H3K4me3 in which both narrow and broad differential peaks are deemed to be differential by epigraHMM at FDR level of 0.05 (see rightmost peaks). ChIPComp, for instance, missed the broad peak on the right while called as differential all narrow peaks on the left. In fact, note that the left most region of Figure 4D is comprised by a super enhancer region, in which several peaks are present in the cell line Hepg2. Our method was able to call this super enhancer region as differential while also covering the broad peak on the right of the panel.

@@

> We believe that Spearman correlation and LFC are not good metrics to compare peak callers with broad marks for two reasons. First, methods that tend to call narrow peaks within broad regions of enrichment will exhibit an unrealistic advantage in performance in terms of LFC. This is so because these methods tend to call as narrow peaks any genomic window exhibiting excessively high read count enrichment (or, LFC in differential enrichment). For broad marks, interest lies not on these single punctate windows but instead in broad regions of enrichment. Second, and due to a similar reason, these methods will also exhibit unrealistic advantage in performance in terms of Spearman correlation because broad regions of enrichment are much noisier than punctate regions and, thefore, read count correlations calculated from broad peaks will be attenuated in comparison to correlations calculated from narrow and punctate windows.

> Regarding the use of 'matched differentially expressed genes as the gold standard to evaluate the differential transcribed genes called from differential H3K36me3 activity', we did not observe a sufficient correspondance between gene expression levels and H3K36me3 enrichment for the cancer cell lines analyzed in our paper. For instance, we have found several instances were genes were expressed but no enrichment of H3K36me3 was observed. Below, we present some of these problematic cases obtained from the UCSC Genome Browser. Note in Figure 1, for example, that both cell lines Helas3 and Hepg2 did not show enrichment for H3K36me3 in the rightmost region of the plot while the associated gene body was clearly expressed. Note in Figure 2, for example, the lack of agreement between the transcription level of the reported gene and the enrichment for H3K36me3 in cell line Hepg2. Due to these reasons, we decided not to use gene expression levels when benchmarking H3K36me3 as suggested by the reviewer.

@@

> Regarding the biological significance of our method, Figure 5E of our paper now presents an analysis of known marker genes for cancer cell line Helas3. Specifically, we show that all known marker genes of Helas3 had associated differential enrichment for the H3K36me3 mixture model component. The presented differential associated enrichment level shown in Figure 5E is the average mixture model posterior proportion associated with the mixture component that model the enrichment of H3K36me3 alone. These results highlight the benefit of the chosen modeling approach that allows not only the detection of differential regions but also the correctly classification of differential enrichment pattern, in this case the gene transcription level associated with H3K36me3.

@@

> Input control samples were utilized by the methods that are designed and capable to do so. These methods include ChIPComp, THOR, diffReps, and DiffBind. We added the following sentence to the second paragraph of Section 5 of our paper: "Results presented in this section pertain to the analysis of two cell lines, namely Helas3 and Hepg2. Analyses were adjusted by input control samples in methods that are designed to do so (ChIPComp, DiffBind, diffReps, and THOR)". Similarly, input controls were also generated in our simulation study and accounted for in these algorithms. Throughout our analysis, we did not observe any significant improvement in performance by accounting for input control effect in differential peak calling. This effect was also reported by others (Lun \& Smyth, 2015).

@@

> We thank the reviewer for the constructive comment. As we replied to the first anonymous reviewer, we did not observe sufficient agreement between gene expression levels for the analyzed cell lines in our paper (see Figures 1 and 2 above). In Figure 2, we have an example of a H3K36me3 differentiallly enriched region and a not DE gene. In contrast, Figure 3 below presents an additional example of a DE gene and without H3K36me3 enrichment in both Helas3 and Hepg2 cell lines. Moreover, not all the analyzed marks in this paper are associated with the expression level of the nearest gene. For instance, H3K27me3 and EZH2 are two epigenomic marks known to exhibit read enrichment in intergenic regions without any nearby gene body. In several instances, enrichment of the repressive mark H3K27me3 occurs on inter-genic regions with nearby expressed genes. Figure 4 below shows an example of such a region. Due to these reasons, we believe that the proposed analysis would be suboptimal to the one presented in this paper for the analyzed cell lines.

@@

> As we discuss above, Figure 5E of our paper now presents an analysis of known marker genes for cancer cell line Helas3. Specifically, we show that all known marker genes of Helas3 had associated differential enrichment for the H3K36me3 mixture model component. The presented differential associated enrichment level shown in Figure 5E is the average mixture model posterior proportion associated with the mixture component that model the enrichment of H3K36me3 alone. These results highlight the benefit of the chosen modeling approach that allows not only the detection of differential regions but also the correctly classification of differential enrichment pattern, in this case the gene transcription level associated with H3K36me3.

>>>>

> We thank the reviewer for this constructive comment. We cite in our paper the benchmarking article 'A comprehensive comparison of tools for differential ChIP-seq analysis' (Briefings in Bioinformatics, 2016). However, we are not aware of a more recent comprehensive review of DPCs algorithms in the literature. As suggested, we add the following sentence to our real data analysis section 'We compared the genome-wide performance of the presented model with the widely used DPCs ChIPComp, csaw, DiffBind, diffReps, RSEG, and THOR, a list that covers a variety of algorithmic approaches for both broad and short ChIP-seq enrichment profiles.' (PEDRO I UPDATED THE PRIOR LINE, PLEASE UPDATE IN THE MAIN TEXT) We believe that such a list covers the most used algorithms for DPC in thie field of bioinformatics.

@@

> As mentioned above, our method is able to account go GC-bias via model offsets. These model offsets can be obtained by methods such as the one proposed by Teng and Irizarry (2017). We present a simulation study in Section B1.3 of the Web Appendix showing that epigraHMM leads to higher sensitivity/specificity when accounting for GC-bias even after read count normalization using the count-normalizing denominators from Teng and Irizarry (2017). However, for the broad marks that we analyzed in our manuscript, we did not observe a substantial effect of GC-content on read count distribution (IF WE EVALUATED AND SHOWED THIS IN THE PAPER, CITE THE SECTION OR FIGURE WHERE WE DID) and did not observe any substantial improvement in accounting for GC-content in our model via offsets when utilizing the method from Teng and Irizarry (2017), a method that was developed only for narrow marks characterized by sharp peaks. In fact, the authors do not recommend the use of their method when analyze epigenomic marks with broad enrichment. For these reasons, we did not explore alternative methods for GC-bias correction for broad marks. (NOTE: SEEMS LIKE YOU ARE SAYING THAT WE TRIED IT FOR BROAD MARKS AND I DID NOT MAKE A BIG DIFFERENCE, AND FOR SHORT MARKS WE CAN USE THE TENG AND IRRIZARY METHOD. DO WE WANT TO SAY WHY IT DOESNT WORK FOR BROAD MARKS HERE?  SIGNAL IS TOO DIFFUSE SO LOWER CORRELATION WITH ENRICHMENT ACROSS WINDOWS COMPARED TO SHORT ENRICHMENT MARKS? I AM ASSUMING IT MADE A DIFFERENCE FOR SHORT MARKS? )

@@

> In the main text, we refer the reader to Section B5 of the Web Appendix (FOR THE FORM OF THE WEIGHTED NB REGRESSION?  IF SO SAY IT TO BE CLEAR). To estimate the model coefficients during the $t^{th}$ iteration of the EM-algorithm, one needs to find the optimal values $\beta_{1}^{(t)}$, $\beta_{3}^{(t)}$, $\lambda_{1}^{(t)}$, and $\lambda_{3}^{(t)}$ that maximize the quantity $Q_{1}\left(\boldsymbol{\psi}_{1}|\boldsymbol{\Psi}^{(t)}\right) + Q_{2}\left(\boldsymbol{\delta},\boldsymbol{\psi}_{2}|\boldsymbol{\Psi}^{(t)}\right) + Q_{3}\left(\boldsymbol{\psi}_{3}|\boldsymbol{\Psi}^{(t)}\right)$. This problem is analogous to estimating parameters in a weighted NB regression model, since $Q_{1}$, $Q_{2}$, and $Q_{3}$ can be seen as weighted log-likelihood functions from a NB generalized linear model with means and dispersions parametrized as described in the Section 3.1 of the main article. Exact forms of the $Q$-functions are presented in Equation B.4 of Web Appendix B5. (I WOULD DIRECTLY EXPLAIN WHAT IS X AND Y HERE AS THE REVIEWER ASKED.  I ALSO DO NOT SEE EXACTLY WHERE YOU EXPLAINED THE EXACT FORM OF THE WEIGHTED NB REGRESSION...IF THAT IS CLEAR FROM SECTION B5 IN THE APPENDIX THEN IGNORE THIS COMMENT).

@@

> Results presented in Table 1 pertain to the full model that utilizes all possible combinatorial patterns in the differential state. This is the 'optimal' model in the sense that the simulated data contains all possible combinatorial patterns of enrichment and background by design in our simulation. The purpose of Figure 1 is to show that our efficient estimation method via fast RCEM leads to unbiased parameter estimates. Together with Figure 2, Table 1 raises the point that increasing the number of technical or biological replicates leads to better results in terms of sensitivity/specificity of peak calls as well as in the sense of asymptotic efficiency of parameter estimates. We leave the discussion of model selection via BIC in our model to Section B2 of the Web Appendix where we present a simulation study that directly evaluates the performance of model selection in our method.

@@

> We thank the reviewer for pointing this out. We believe that comparing the relative bias of parameter estimates across different SNR levels is difficult because different SNRs are associated with different parameter values (some larger than others) in our simulation study. Although we did observe an improvement in performance in terms of model sensitivity/specificity of peak calls in higher SNR data sets, the main driving quantity for the reduction of parameter bias is the number of replicates included in our model. For this reason, we rephrase the mentioned sentence in our paper as "Depending on the number of conditions, the observed relative bias and the range of the reported percentiles tended to decrease as more replicates were included in the analyses. This effect was particularly significant in scenarios with four conditions with respect to parameters $\beta_{3}$ and $\lambda_{3}$. In scenarios with higher number of conditions, these results highlight the importance of experimental replicates to achieve precise parameter estimates.". The estimated SNR level from ENCODE was 3.21 (ratio of mean counts between peak and background regions) for H3K27me3. This is now reported in our manuscript. We present the estimate SNR value for other scenarios in our Web Appendix (Web Tables 2-3)

@@

> We believe our method is able to handle such situations. Note, for instance, in Figure 4D of our manuscript that we report differential peak calls from H3K4me3 in which both narrow and broad differential peaks are deemed to be differential by epigraHMM at an FDR level of 0.05 (see rightmost peaks). ChIPComp, for instance, missed the broad peak on the right while called as differential all narrow peaks on the left. In fact, note that the left most region of Figure 4D is comprised by a super enhancer region, in which several peaks are present in the cell line Hepg2. Our method was able to call this super enhancer region as differential while also covering the broad peak on the right of the panel.

@@

> We believe that Spearman correlation and LFC are not good metrics to compare peak callers with broad marks for two reasons. First, methods that tend to incorrectly call narrow peaks within broad regions of enrichment will exhibit inflated performance in terms of LFC, despite not capturing broad signal regions. This is because these methods tend to call as narrow peaks any genomic window exhibiting excessively high read count enrichment (or, LFC in differential enrichment) (IT WOULD BE GOOD TO POINT IN AN EXAMPLE IN A FIGURE FROM THE PAPER THAT SHOWS AN EXAMPLE). For broad marks, interest lies not on these single punctate windows but instead in broad regions of enrichment. Similarly, methods that accurately call broad regions of enrichment capture signal are much noisier than in punctate regions (SHOW EXAMPLE OF THIS) and, thefore, read count correlations calculated from broad peaks will be attenuated in comparison to correlations calculated from those called from narrow and punctate windows.

> Regarding the use of 'matched differentially expressed genes as the gold standard to evaluate the differential transcribed genes called from differential H3K36me3 activity', we did not observe a sufficient correspondance between gene expression levels and H3K36me3 enrichment for the cancer cell lines analyzed in our paper. For instance, we have found several instances where genes were expressed but no enrichment of H3K36me3 was observed. Below, we present some of these problematic cases obtained from the UCSC Genome Browser. Note in Figure 1, for example, that both cell lines Helas3 and Hepg2 did not show enrichment for H3K36me3 in the rightmost region of the plot while the associated gene body was clearly expressed. Note in Figure 2, for example, the lack of agreement between the transcription level of the reported gene and the enrichment for H3K36me3 in cell line Hepg2. Due to these reasons, we decided not to use gene expression levels when benchmarking H3K36me3 as suggested by the reviewer. (I AM STILL NOT FOLLOWING THIS ARGUMENT.  HOW ARE WE BENCHMARKING THEN?)

@@

> Regarding the biological significance of our method, Figure 5E of our paper now presents an analysis of known marker genes for cancer cell line Helas3. Specifically, we show that all known marker genes of Helas3 were associated with differential enrichment for the H3K36me3 mixture model component (I DONT NKOW WHAT THIS MEANS - WINDOWS THAT OVERLAP THESE GENES WERE CLASSIFIED AS THE DIFFERENTIAL STATE?  MAKE THIS CLEAR). The presented differential associated enrichment level shown in Figure 5E is the average mixture model posterior proportion associated with the mixture component that model the enrichment of H3K36me3 alone (THIS LINE IS HARD TO FOLLOW...DO YOU MEAN THE MEAN K36 ONLY COMBINATION POSTERIOR PROBABILITY OF WINDOWS COVERING THIS GENE?  NOT CLEAR). These results highlight the benefit of the chosen modeling approach that allows not only the detection of differential regions but also the correctly classification of differential enrichment pattern, in this case the gene transcription level associated with H3K36me3.

@@

> Input control samples were utilized by the methods that are designed and capable to do so. These methods include ChIPComp, THOR, diffReps, and DiffBind. We added the following sentence to the second paragraph of Section 5 of our paper: "Results presented in this section pertain to the analysis of two cell lines, namely Helas3 and Hepg2. Analyses were adjusted by input control samples in methods that are designed to do so (ChIPComp, DiffBind, diffReps, and THOR)". Similarly, input controls were also generated in our simulation study and accounted for in these algorithms. Throughout our analysis, we did not observe any significant improvement in performance by accounting for input control effect in differential peak calling (IS THIS REFERRING TO ALL METHODS, OR JUST OURS?  DO WE SHOW THIS FOR EITHER?  IF SO WE SHOLD REFER TO THE FIGURE OR SECTION RELATED TO THIS POINT). This effect was also reported by others (Lun \& Smyth, 2015).

@@

> We thank the reviewer for the constructive comment. As we replied to the first anonymous reviewer, we did not observe sufficient agreement between gene expression levels for the analyzed cell lines in our paper (see Figures 1 and 2 above). In Figure 2, we have an example of a H3K36me3 differentiallly enriched region and a not DE gene. In contrast, Figure 3 below presents an additional example of a DE gene and without H3K36me3 enrichment in both Helas3 and Hepg2 cell lines. Moreover, not all the analyzed marks in this paper are associated with the expression level of the nearest gene. For instance, H3K27me3 and EZH2 are two epigenomic marks known to exhibit read enrichment in intergenic regions without any nearby gene body. In several instances, enrichment of the repressive mark H3K27me3 occurs on inter-genic regions with nearby expressed genes. Figure 4 below shows an example of such a region. Due to these reasons, we believe that the proposed analysis would be suboptimal to the one presented in this paper for the analyzed cell lines. (SO THE ARGUMENT IS THAT IT ISNT A 1:1 RELATIONSHIP AND THATS WHY WE DIDNT USE DIFFERENTIAL EXPRESSION OF GE WITH DIFFERENTIAL EXPRESSION OF THE MARK?  HOW WIDESPREAD IS THIS?  DO A LOT OF OTHER PAPERS DO THIS?  IF NOT, CAN WE CITE OTHER PAPERS ALSO, SAYING THAT THEY DO NOT USE THIS METRIC?)

@@

> As we discuss above, Figure 5E of our paper now presents an analysis of known marker genes for cancer cell line Helas3. Specifically, we show that all known marker genes of Helas3 had associated differential enrichment for the H3K36me3 mixture model component. The presented differential associated enrichment level shown in Figure 5E is the average mixture model posterior proportion associated with the mixture component that model the enrichment of H3K36me3 alone (SEE MY PRIOR COMMENT ABOUT MAKING THIS CLEAR...POSTERIOR PROBABILITY OF K36 ALONE, AVERAGED ACROSS WINDOWS SPANNING THIS GENE?). These results highlight the benefit of the chosen modeling approach that allows not only the detection of differential regions but also the correctly classification of differential enrichment pattern, in this case the gene transcription level associated with H3K36me3.
